---
output: github_document
---
[![codecov](https://codecov.io/gh/rmsharp/nprcmanager/branch/master/graph/badge.svg)](https://codecov.io/gh/rmsharp/nprcmanager)

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# nprcmanager
## Introduction
The goal of nprcmanager is implement Genetic Tools for Colony Management.
It was initially conceived and developed at
the Oregon National Primate Research Center (ONPRC) to facilitate some of
the analyses they perform regularly.

At present, the application is designed to support 3 functions:

    1. Quality control of uploaded studbooks
    2. Generation of Genetic Value Analysis Reports
    3. Creation of potential breeding groups
    4. Display of an age by sex pyramid plot


## Installation

You can install nprcmanager from github with:

```{r gh-installation, eval = FALSE}
install.packages("devtools")
devtools::install_github("rmsharp/nprcmanager")
```

## Running the Shiny application in nprcmanager

The toolset available within nprcmanager can be used inside standard R scripts.
However, it was orginally designed to be used within a Shiny application 
that can be started with:

```{r example, eval = FALSE}
library(nprcmanager)
runManager()
```
## Summary of Major Functions of Shiny Application

### Quality Control
Studbooks maintained by breeding colonies generally contain
information of varying quality. The quality control functions
of the toolkit check to ensure all animals listed as parents
have their own line entries, all parents have
the appropriate sex listed, no animals are
listed as both a sire and a dam, duplicate entries are removed,
and all dates are valid dates. 
In addition, exit dates are
added if possible and are consistant with other information such as 
departure dates and death dates, parents with ages below a user selected
threshold are identified, pedigree generation numbers are added, current
ages of animals that are still alive are added, .

Further quality control measures may be added later, such as
checking to ensure parents' birthdates precede their children's.

### Genetic Value Analysis Reports
The Genetic Value Analysis is a ranking scheme developed at ONPRC to
indicate the relative breeding value of animals in the colony. The
scheme uses the mean kinship for each animal to indicate how
inter-related it is with the rest of the current breeding colony
members. Genome uniqueness is used to provide an indication of
whether or not an animal is likely to possess alleles at risk
of being lost from the colony. Under the scheme, animals with low
mean kinship or high genome uniqueness are ranked more highly.

### Breeding Group Formation
One of the goals in breeding group formation is to avoid the
potential for mating of closely related animals. Since behavioral
concerns and housing constraints will also be taken into account
in the group formation process, it is our goal to provide the largest
number of animals possible from a list of candidates that can be housed
together without risk of consanguineous mating. To that end, this
function uses information from the Genetic Value Analysis to search
for the largest combinations of animals that can be produced from
a list of candidates.


**For more information see:**  
A Practical Approach for Designing Breeding Groups to Maximize Genetic 
Diversity in a Large Colony of Captive Rhesus Macaques (*Macaca mulatto*)
Vinson, A ; Raboin, Mj
*Journal Of The American Association For Laboratory Animal Science*, 
2015 Nov, Vol.54(6), pp.700-707 [Peer Reviewed Journal]

## Use with LabKey
Two configuration files are needed to use the database features of nprcmanager
with LabKey. 
The first file is named **\_netrc** on Microsoft Windows operating systems and 
**.netrc** otherwise, allows the user to authenticate with LabKey through the 
LabKey API and is fully 
described by [LabKey documentation](https://www.labkey.org/Documentation/wiki-page.view?name=netrc)

The second file is named **\_nprcmanager_config** on Microsoft Windows 
operating systems and **.nprcmanager_config** otherwise and is the nprcmanager 
[configuration file](https://github.com/rmsharp/nprcmanager/blob/master/inst/extdata/example_nprcmanager_config)
