---
output: github_document
---
[![Rdoc](http://www.rdocumentation.org/badges/version/roxygen2)](http://www.rdocumentation.org/packages/roxygen2)
[![Build Status](https://travis-ci.org/rmsharp/nprcmanager.svg?branch=master)](https://travis-ci.org/rmsharp/nprcmanager)
[![codecov](https://codecov.io/gh/rmsharp/nprcmanager/branch/master/graph/badge.svg)](https://codecov.io/gh/rmsharp/nprcmanager)
<!--[![Rdoc](http://www.rdocumentation.org/badges/version/RDocumentation)](http://www.rdocumentation.org/packages/RDocumentation)
<!--[![Rdoc](http://www.rdocumentation.org/badges/version/nprcmanager)](http://www.rdocumentation.org/packages/gh/rmsharp/nprcmanager)
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# nprcmanager -- Version `r nprcmanager::getVersion()`
```{r child = "vignettes/manual_components/_introduction.Rmd", ref = "introduction"}
```
```{r child = "vignettes/manual_components/_installation.Rmd", ref = "installation"}
```
```{r child = "vignettes/manual_components/_summary_of_major_functions.Rmd", ref = "summary_of_functions"}
```
```{r child = "vignettes/manual_components/_running_shiny_application.Rmd", ref = "running_shiny_application"}
```

Find online documentation at [https://rmsharp.github.io/nprcmanager/](https://rmsharp.github.io/nprcmanager/).


**For more information see:**  
A Practical Approach for Designing Breeding Groups to Maximize Genetic 
Diversity in a Large Colony of Captive Rhesus Macaques (*Macaca mulatto*)
Vinson, A ; Raboin, MJ
*Journal Of The American Association For Laboratory Animal Science*, 
2015 Nov, Vol.54(6), pp.700-707 [Peer Reviewed Journal]

