---
title: "Genetic Management Tools"
author: "Michael Raboin and R. Mark Sharp"
date: "October 03, 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Genetic Management Tools}
  %\usepackage[UTF-8]{inputenc}
---
---

<a href="#introduction">Introduction</a>  
<a href="#summary_of_functions">Summary of Functions</a>  
<a href="#installation">Installation</a>  
<a href="#running_shiny_application">Running Shiny Application</a>  
<a href="#input">Input</a>  
<a href="#pedigree_browser">Pedigree Browser</a>  
<a href="#gva">Genetic Value Analysis</a>  
<a href="#summary_statistics">Summary Statistics</a>  
<a href="#breeding_group_formation">Breeding Group Formation</a>  
<a href="#orip">ORIP Reporting</a>  
<a href="#readme">README</a>  
<a href="#breeding_group_algorithm">Algorithm: Breeding Group Formation</a>  
<a href="#genome_uniqueness_algorithm">Algorithm: Genome Uniqueness</a>  

```{r child = "introduction.Rmd", ref = "introduction"}
```
```{r child = "summary_of_major_functions.Rmd", ref = "summary_of_functions"}
```
```{r child = "installation.Rmd", ref = "installation"}
```
```{r child = "running_shiny_application.Rmd", ref = "running_shiny_application"}
```
```{r child = "input.Rmd", ref = "input"}
```
```{r child = "pedigree_browser.Rmd", ref = "pedigree_browser"}
```
```{r child = "genetic_value_analysis.Rmd", ref = "gva"}
```
```{r child = "summary_statistics.Rmd", ref = "summary_statistics"}
```
```{r child = "breeding_group_formation.Rmd", ref = "breeding_group_formation"}
```
```{r child = "gv_and_bg_desc.Rmd", ref = "gv_and_bg_desc"}
```
```{r child = "breeding_group_algorithm.Rmd", ref = "breeding_group_algorithm"}
```
```{r child = "genome_uniqueness_algorithm.Rmd", ref = "genome_uniqueness_algorithm"}
```

<h2 id="orip">ORIP Reporting</h2>
The ORIP Reporting tab will eventually contain information for reporting to the
Office of Research Infrastructure Programs (ORIP). This tab may end up being merged
with the Summary Statistics tab and contain a number of statistics, tables and
histograms. Alternatively, this may contain a subset of information from the
Summary Statistics tab presented as a formatted report that can be exported and
submitted to ORIP. The exact information that needs to be submitted for ORIP
recordkeeping is still under discussion.

