---
title:
  formatted: Genetic Colony Management with \pkg{nprcmanager}
  plain: Genetic Colony Management with nprcmanager
  short: '\pkg{nprcmanager}: Genetic Colony Management'
author:
- name: R. Mark Sharp
  affiliation: Deppe
  address: |
    7526 Meadow Green St San Antonio, TX 78251
  email: \email{rmsharp@me.com}
  url: http://rmsharp.me.com
- name: Amanda Vinson
  affiliation: |
    Oregon Health and Science University  ONPRC
abstract: |
  The abstract of the article.
output:
  pdf_document: default
  html_document: default
preamble: |
  \usepackage{amsmath}
keywords:
  formatted:
  - genetic
  - nonhuman
  - \proglang{Shiny}
  - \proglang{R}
  - management
  - captive
  - population
  - inbreeding
  - kinship
  - simulation
  plain:
  - genetic
  - nonhuman
  - Shiny
  - R
  - management
  - captive
  - population
  - inbreeding
  - kinship
  - simulation
vignette: |
  %\VignetteEngine{knitr::knitr} %\VignetteIndexEntry{Interactive Use} %\usepackage[UTF-8]{inputenc}
---
```{r setup}
library(shiny)

```

# Introduction

This tutorial provides examples of the functions provided within the \pkg{nprcmanager} package.


## Code formatting

Don't use markdown, instead use the more precise latex commands:

* \proglang{Java}
* \pkg{plyr}
* \code{print("abc")}

# R code

Can be inserted in regular R markdown blocks.

```{r}
x <- 1:10
x
```

# Importing in the Pedigree Data

## Pedigree File Structure


It begins with importing a simple pedigree file, pedigree file with genotypes, pedigree file and
and associated set of genotypes, and a initiating a database query that
returns the full pedigree associated with a list of animals provided by the user.


The recommended first step after creation of an input file is checking  the file for
detectable errors.

## Selecting the Minimam Parental Age

## Pedigree File Only

The field seperator used within a pedigree file must be the same throughout
and must be indicated when reading in the file. 
Fields can be separated by tabs, commas, or simicolons.


## Pedigree and Genotypes in the Same File

## Pedigree and Genotypes in Separate Files

## Pedigree Provided by LabKey Query

